% preamble

\input{preamble}

% document

\begin{document}

\section*{$p-$adic valuations and digit sums \begin{small} \begin{color}{gray} \\
(\today) \end{color} \end{small}}

Previous posts have discussed the prime factorizations of the central binomial coefficients, $\ds \gamma_n =\binom{2n}{n}$, arriving at the question of the $p-$adic valuation of $\ds \gamma_n, v_p(\gamma_n)$, for relatively small primes $\ds (< 2n/3)$. There is an interesting relationship between $p-$adic valuations and digit sums. We typically write numbers in base ten, so the symbol $\ds 352$ abbreviates the number
\[
  3 \times 10^2 + 5\times 10^1 + 2\times 10^0
\]
Any positive integer $p$ can be used as a base. Given a positive integer $n$, the base $p$ digits of $n$ are numbers $\ds d_0, d_1, \ldots, d_m$ with $\ds 0\leq d_i <p$ such that
\[
  n = d_0 p^0 + d_1 p^1 + \cdots d_m p^m
\]
For example 
\[
  352 = 2\times 5^3 + 4\times 5^2 + 0\times 5^1 + 2\times 5^0
\]
so the base $5$ digits of the number $352$ are $\ds d_0 = 2, d_1 = 0, d_2=4, d_3=2$. We'll use the notation $\ds S_p(n)$ to express the sum of the base $p$ digits of $n$, so $\ds S_5(352) = 2+4+0+2=8$ and $\ds S_{10}(352) = 3+5+2=10$.

If $\ds v_p(k)=l>0$, then the base $p$ representation of $k$ has zeroes in the rightmost $l$ places in the same way that a number divisible by 1000 ends in three zeroes (in base ten). That is, $\ds k = d_l p^l + d_{l+1}p^{l+1} + \cdots + d_m p^m$. This means: 
\begin{eqnarray*}
 k-1 & = & (p^l-1) + (d_l-1)p^l+ d_{l+1}p^{l+1} + \cdots + d_m p^m \\
   & = & \sum_{i=0}^{l-1} (p-1)p^i + (d_l-1)p^l+ d_{l+1}p^{l+1} + \cdots + d_m p^m \\
\end{eqnarray*}
So if we add up the digits above
\[
  S_p(k-1) = (p-1)l + S_p(k) - 1
\]
or
\begin{equation} \label{PreLegendre}
  v_p(k) = \frac{1 + S_p(k-1) - S_p(k)}{p-1}
\end{equation}
If $\ds v_p(k)=0$, then the ones digit of $k$ (in base $p)$ is not zero and so the base $p$ digits of $k-1$ are identical to those of $k$ except the ones digit is one less, meaning $\ds S_p(k-1) = S_p(k) -1$, and (\ref{PreLegendre}) still holds (both sides of the equation are zero).

What is interesting about this result is that it gives a very concrete relationship between two somewhat distinct aspects of a number--its $p-$adic valuation and its base $p$ digit sum.


\iffalse
\section{A Formula from Legendre}
\label{AFormulaFromLegendre}

Prime factorizations of products concatenate with products of powers of the same base simplifying according to the rules of exponents. For example, the prime factorization of $9450$ is $\ds 2\cdot 3^3 \cdot 5^2\cdot 7$ and the prime factorization of $\ds 7938$ is $\ds 2\cdot 3^4\cdot 7^2$, so the prime factorization of $\ds (9450)(7938)$ is 
\[
  \lp 2\cdot 3^3 \cdot 5^2\cdot 7 \rp \lp 2\cdot 3^4\cdot 7^2 \rp = 2^2 \cdot 3^7\cdot 5^2\cdot 7^3
\]
Notice that $\ds v_3\lp (9450)(7938) \rp = v_3(9450) + v_3(7938)$. The $p-$adic valuation function has the log-like property that 
\[
  v_p(ab) = v_p(a) + v_p(b)
\]

Consequently
\[
  v_p(n!) = v_p(n(n-1)\cdots 2\cdot 1) = \sum_{k=1}^n v_p(k)
\]
Using the relationship between $p-$adic valuations and digit sums discussed in Section \ref{PAdicValuationsAndDigitSums}, we can re-express this:
\begin{eqnarray*}
  v_p(n!) & = & \sum_{k=1}^n \frac{1 + S_p(k-1) - S_p(k)}{p-1} \\
  & = & \frac{1}{p-1}\lp n + S_p(0) - S_p(n) \rp \\
  & = & \frac{n - S_p(n)}{p-1}
\end{eqnarray*}
This result was proved by Legendre in 1830.

\section*{Kummer's Theorem}
\label{KummersTheorem}

In Section \ref{PowersOfTwo}, we considered the $2-$adic valuations of the central binomial coefficients $\ds \gamma_n = \binom{2n}{n}$, which numbers are symbolized by $\ds v_2(\gamma_n)$. Playing around with this function, one is struck by the observation that $\ds n = 2^k$ for some nonnegative integer $k$ if and only if $\ds v_2\lp \gamma_n \rp = 1$. This is a consequence of Kummer's Theorem, which says that if $\ds S_p(x)$ is the sum of the base $p$ digits of $x$, then
\begin{equation} \label{kummer}
  v_p\lp \binom{r}{k}\rp = \frac{S_p(k) + S_p(r-k) - S_p(r)}{p-1}
\end{equation}
Because of the log-like nature of the $p-$adic valuation mentioned in Section \ref{AFormulaFromLegendre}, 
\[
  v_p\lp \binom{r}{k}\rp = v_p \lp \frac{r!}{k!(r-k)!} \rp = v_p(r!) - v_p(k!) - v_p((r-k)!)
\]
Using Legendre's Formula from that same section to replace the terms on the right and simplifying gives (\ref{kummer}), and applying (\ref{kummer}) to the central binomial coefficients we get 
\begin{equation} \label{VpOfGamman}
  v_p(\gamma_n) = \frac{2 S_p(n) - S_p(2n)}{p-1}
\end{equation}

Multiplying a number by 10 shifts its base ten digits one place to the right but does not change its digit sum. More generally, multiplying a number by $p$ does not change its base $p$ digit sum. So in particular, $\ds S_2(2n) = S_2(n)$, and in the case $\ds p=2$, (\ref{VpOfGamman}) becomes the remarkable formula
\[
  v_2(\gamma_n) = S_2(n)
\]
That is, the power of $2$ that appears in the prime factorization of $\ds \binom{2n}{n}$ is just the sum of the base two digits of $n$, also called the \textbf{Hamming Weight} of $n$.

If the base $p$ digits of $n$ are $\ds d_0, d_1, \ldots, d_k$, and $\ds d_i \leq (p-1)/2$ for all $i$, then the base $p$ digits of $2n$ are $\ds 2d_0, 2d_1, \ldots, 2d_k$, which means that $\ds S_p(2n) = 2 S_p(n)$ and so $\ds v_p(\gamma_n)=0$. For example, $\ds 27=220_5$, so it should be the case that $\ds v_5(\gamma_{27}) =0$ (and it is). As $\ds v_p(\gamma_n)=0$ if and only if $\ds S_p(2n) = 2 S_p(n)$, the primes that do not appear in the prime factorization of $\ds \gamma_n$ are precisely those primes $p$ such that none of the base $p$ digits of $n$ is more than $\ds (p-1)/2$.

Let $q = (p-1)/2$ for a prime $p>3$, and let $\ds \lb \gamma_{n_k} \rb$ be the subsequence of $\ds \lb \gamma_n \rb$ such that $\ds v_p(\gamma_{n_k})=0$. $\ds n_k$ is the number whose base $p$ digits are the base $q$ digits of $k$. For example, suppose we want to find the $\ds 147^{th}$ positive integer $\alpha$ such that $\ds v_{13}(\gamma_{\alpha})=0$. First we express $\ds 147$ in base $\ds (13-1)/2=6$: $\ds 147 = 403_6$. Then $\ds \alpha$ is the number with this base 13 representation, so 
\[
  \alpha = 4\times 13^2 + 0\times 13 + 3\times 13 = 679
\]
and one can check that $\ds v_{13}(\gamma_{679}) = 0$.

Conjecture: $\ds v_p(\gamma_n) = 1$ if and only if the set of base $p$ digits of $n$ has exactly one element that is larger than $\ds (p-1)/2$.



\iffalse
A few observations that will hopefully lead to a proof of this useful theorem. First,  
\[
  v_p(\gamma_n) = v_p( (2n)!) - 2 v_p(n!)
\]

This implies that $\ds v_p\lp \gamma_n \rp = \frac{2 S_p(n) - S_p(2n)}{p-1}$. $\ds S_2(2n) = S_2(n)$, and if $n$ is a power of 2, then $\ds S_2(n)=1$ and so $\ds v_2 \lp \gamma_n \rp = 1$.

A consequence of this, as $\ds v_p(\gamma_n)$ must be a nonnegative integer, is that for any prime $p$, $\ds 2 S_p(n) - S_p(2n)$ must be divisible by $\ds p-1$.


\fi



\section*{Erd\"{o}s Conjecture}

Erd\"{o}s conjectured that that for $\ds n>4$, $\ds \gamma_n$ is not squarefree, which is equivalent to saying that for $n>4$, there is some prime $p$ such that $\ds p^2$ divides $\ds \gamma_n$. As we have seen, any such prime must be fairly small, i.e., must come from the range $\ds \lb 1,2,\ldots, 2n/3\rb$. What's more, a consequence of the observation that $\ds v_2(\gamma_n) = S_2(n)$ made in Section \ref{KummersTheorem} is that if $\ds S_2(n)>1$, then $\ds \gamma_n$ is divisible by four and so not square free. The only positive integers $n$ with $\ds S_2(n)\leq 1$ are powers of two, so if there is a positive integer $\ds n$ such that $\ds \gamma_n$ is not squarefree, then $n$ must be a power of two.

For a positive integer $n$, $n$ \textbf{primorial} is the product of the primes less than or equal to $n$ and is expressed $\ds n\# $. So
\begin{eqnarray*}
 10\# & = & (2)(3)(5)(7) = 210 \\
 15 \# & = & (2)(3)(5)(7)(11)(13) = 30030
\end{eqnarray*}
The prime factorization of $\ds \gamma_n$ includes one copy of every prime in the range $\ds \lb n+1, n+2, \ldots, 2n\rb$. The product of such primes is $\ds \frac{(2n)\#}{n \#}$. It can be shown that $\ds \frac{(2n)\#}{n \#} \leq 3.439^n$.



The number on the right of (\ref{FactoringGamma50}) is the product of ten consecutive primes, from the $\ds 16^{th}$ prime through the $\ds 25^{th}$ prime. 
So the number in (\ref{FactoringGamma50}) is $\ds \frac{97 \#}{52 \#}$.



\begin{thebibliography}{99} 

\bibitem{HANSON} Hanson, Denis. ``On the Product of the Primes." Canadian Mathematical Bulletin, volume 15, issue 1, March, 1972, pp. 33-37.

\end{thebibliography}
\fi


\end{document}
